Local PCA Shows How Population Structure Differs Along the Genome

### 2016-07-09 writeup/local_pca_paper.tex
done editing

The general steps in our method could be carried out in many ways.
Referring to Figure \ref{fig:diagram}, these are
(1) divide the genome into windows,
(2) summarize the patterns of relatedness (population structure) in each window,
(3) measure dissimilarity in population structure between each pair of windows,
(4) visualize the resulting dissimilarity matrix,
and (5) combine similar windows to more accurately estimate local population structure.
Details of how we carried these out are given below.

@@

We used three publically available datasets (summarized in Table \ref{tab:data_stats});
as usual for genetic data we converted the data to a numeric matrix
(with one row per polymorphic variant and one column per sample)
by replacing each genotype with the number of nonreference alleles
(or NA for missing data).
A normalization step (see below) ensures the result does not depend on the choice of reference allele.

\paragraph{Human:}
We used genomic data from the entire POPRES dataset \citep{nelson2008population},
which has array-derived genotype information for 447,267 SNPs across the autosomes
of 3,965 samples in total: 346 African-Americas, 73 Asians, 3,187 Europeans and 359 Indian Asians.
(We excluded the sex chromosomes and the mitochondria.)
We use the allele that has highest frequency in the samples as the reference allele for each position.

\paragraph{\textit{Drosophila melanogaster}:}
We also used whole-genome sequencing data 
from the Drosophila Genome Nexus \citep[\url{http://www.johnpool.net/genomes.html}][]{lack2015drosophila},
consisting of the Drosophila Population Genomics Project phases 1--3 \citep[DPGP][]{langley2012genomic,pool2012population},
and additional African genomes \citep[AGES][]{lack2015drosophila},
After removing 20 genomes with more than 8\% missing data,
we were left with 380 samples from 16 countries across Africa and Europe.
Since the \textit{Drosophila} samples are from inbred lines, we treat the samples as haploid when recoding;
regions with residual heterozygosity were marked as missing in the original dataset;
we also removed positions with more than 20\% missing data. 
% (We chose these cutoffs as the tails of the relevant empirical distributions.)
Each chromosome arm we investigated (X, 2L, 2R, 3L, and 3R) has 2--3 million SNPs.

@@

\paragraph{\textit{Medicago truncatula}:}
Finally, we used whole-genome sequencing data from the \textit{Medicago truncatula} Hapmap Project \citep{tang2014improved},
which has 263 samples from 24 countries,
primarily distributed around the Mediterranean basin.
Each of the 8 chromosomes has 3-5 million SNPs;
we did not use the mitochondria or chloroplasts.

\begin{table}[ht]
\centering
    \begin{tabular}{p{0.8in}rrrr}
  \hline
    species 
    & \parbox[t]{.8in}{\# SNPs per \\ window} 
    & \parbox[t]{1in}{mean window\\ length (bp)}
    & \parbox[t]{1.2in}{mean \# windows \\ per chromosome} 
    & \parbox[t]{1.4in}{mean \% variance ex-\\plained by top 2 PCs} \\ 
  \hline
  Human & 100 & 636,494 & 203 & 0.55 \\ 
  \textit{Drosophila melanogaster} & 1,000 & 9,019 & 2,674 & 0.53 \\ 
  \textit{Medicago truncatula} & 10,000 & 102,580 & 467 & 0.50 \\ 
   \hline
\end{tabular}
\caption{
    Descriptive statistics for each dataset used,
    with the chosen window sizes.
    See text for how window sizes were chosen;
    in each window we performed PCA,
    computed the percent varince explained by the top two principal components;
    the mean across windows is given in the last column.
    \label{tab:data_stats}
}
\end{table}

@@

After recoding, we divided the genome into contiguous segments (``windows''; see below),
% recoded matrix into contigous matrices that have the same columns but fewer rows than the original matrix, 
and applied Principal Component Analysis (PCA) as described in \citet{mcvean2009genealogical}
separately to the submatrices that corresponded to each window.
Specifically, we did PCA as follows:
denote by $Z$ the $L\times N$ recoded genotype matrix ($L$ is the number of SNPs and $N$ is the sample size), 
and by $\overline{Z_{s}}$ the mean of non-missing entries for allele $s$, 
so that $\overline{Z_{s}}=\frac{1}{n_s}\sum_j Z_{sj}$, 
where the sum is over the $n_s$ nonmissing genotypes.
We first compute the mean-centered matrix $X$, as $X_{si}=Z_{si}-\overline{Z_{s}}$,
and preserving missingness.
(This mean-centering makes the result not depend on the choice of reference allele,
exactly if there is no missing data, and approximately otherwise.)
Next, we find the covariance matrix of $X$, denoted $C$,
as $C_{ij} = \frac{1}{m_{ij}-1} \sum_s X_{si} X_{sj} - \frac{1}{m_{ij}(m_{ij}-1)} (\sum_s X_{si})(\sum_s X_{sj})$,
where all sums are over the $m_{ij}$ sites where both sample $i$ and sample $j$ have nonmissing genotypes.
% We compute the covariance matrix using the R function cov() , with use=``pairwise",
% which computes the covariance between each pair of individuals using all complete pairs of SNPs on those individuals.
The principal components are the eigenvectors of $C$, 
normalized to have Euclidean length equal to one,
and ordered by magnitude of the eigenvalues.

@@

Since eigenvectors are still only defined up to sign,
when comparing between windows we choose the sign to best match each other:
after choosing $\pcone_1$, for instance, and $u$ is the first eigenvector obtained from the covariance matrix
for window $j$,
then $\pcone_j = \pm u$,
where the sign is chosen according to which of 
$\sum_i (\pcone_{i1} - u_i)^{2}$ or
$\sum_i (\pcone_{i1} + u_i)^{2}$ 
is smaller.

@@

The choice of window length entails a balance between several factors.
In very short windows, genealogies of the samples will only be represented by a few trees,
so variation between windows represents demographic noise rather than meaningful variation in population structure.
Longer windows generally have more distinct trees (and SNPs), 
allowing for less noisy estimation of local population structure.
However, to better resolve meaningful signal, i.e., differences in population structure along the genome, 
we would like reasonably short windows.
Window length choice therefore entails a signal versus noise tradeoff
in the estimates of population structure.
% If we use the first principal component as a measure of population structure, 
% then to choose the best window length,
% we need to find a balance between the standard error of the first principal components for each window 
% and the standard deviation between windows. 
Since we summarize population structure using relative positions in the principal component maps,
we quantify ``noise'' as the standard error of a sample's position on PC1 in a particular window,
averaged across windows and samples,
and ``signal'' as the standard deviation of the sample's position on PC1 over all windows,
averaged over samples.
(Recall that the signs for PCs are chosen to match each other.)
Then, the mean variance across windows is
\begin{align*}
    \sigma_\text{signal}^2
    = 
    \frac{1}{N} \sum_{j=1}^{N}
        \frac{1}{L}\sum_{i=1}^{L}\left ( \pcone_{ij} -\overline\pcone_{j} \right )^{2} ,
\end{align*}
where $\overline\pcone_j = (1/N) \sum_{j=1}^N \pcone_{ij}$.
We estimate the standard error for each $\pcone_{ij}$ using the block jackknife \citep{efron1982jackknife,busing1999deletem}:
we divide the $j^\text{th}$ block into 10 equal-sized pieces,
and let $\pcone_{ij,k}$ denote the first principal component of this region found after removing the $k^\text{th}$ piece;
then the estimate of the squared standard error is
$\sigma^2_{ij} = \frac{9}{10} \sum_{k=1}^{10} ( \pcone_{ij,k} - \frac{1}{10} \sum_{\ell=1}^{10} \pcone_{ij,\ell} )^2$,
and we measure noise by averaging over samples and windows:
\begin{align*}
    \sigma^2_\text{noise}
    &=
    \frac{1}{N} \sum_{j=1}^{N} \frac{1}{L}\sum_{i=1}^{L} \sigma^2_{ij} .
\end{align*}

We calculated $\sigma^2_\text{signal}$ and $\sigma^2_\text{noise}$
for a range of window sizes,
and for our main results
chose the smallest window for which $\sigma^2_\text{signal}$ was consistently larger than $\sigma^2_\text{noise}$;
the values for various window sizes across \textit{Drosophila} chromosomes are shown in Table \ref{tab:window_sizes},
and the choices for all taxa are in Table \ref{tab:data_stats}.
In practice, even substantially different window sizes result in the same genome-wide patterns.
We also chose windows to each consist of the same number of neighboring SNPs.
We compared to the results when windows are each of the same physical length (in bp) but of the same average length,
and results were nearly identical.

@@

We think of local population structure as being summarized by \emph{relative} position of the sampmles
in the space defined by the top principal components.
Since rotations of this space imply identical population structure,
we do not compare population structure of different genomic windows using the PCs directly, 
but instead compare the low-dimensional approximations of the local covariance matrices
obtained using the top $k$ PCs.
(For results shown here, we use $k=2$;
results using larger numbers of PCs were nearly identical.)
We do this, rather than using the entire covariance matrix, 
both for computational efficiency 
and because the top PCs summarize important population structure, 
so using only these should reduce the effect of noise.
To remove the effect of variation in mutation rate between windows,
we also rescale each covariance matrix so that the underlying data matrix has trace norm equal to one.
To do this, define the $N \times k$ matrix $V(i)$ so that the $\ell^\text{th}$ column of $V(i)$
is equal to the $\ell^\text{th}$ princpal component of the $i^\text{th}$ window,
multiplied by $\sqrt{ \lambda_{\ell i} / \sum_{m=1}^k \lambda_{m i} }$.
% \begin{equation}
%     \begin{aligned}
%         V_{i1}&=\sqrt{\frac{\lambda _{1i}}{\lambda _{1i}+\lambda _{2i}}}PC1_{i} ,
%         \qquad
%         V_{i2}&=\sqrt{\frac{\lambda _{2i}}{\lambda _{1i}+\lambda _{2i}}}PC2_{i} 
%     \end{aligned}
% \end{equation}
Then, the rescaled, $k$-dimensional covariance matrix for the $i^\text{th}$ window is
\begin{align} \label{eqn:est_cov}
    M(i) &= \sum_{\ell=1}^k V_{i\ell} V_{i\ell}^T .
\end{align}
% \begin{align}
%     M_{i} &= \frac{\lambda_{1i}PC1_{i}PC1_{i}^{T}+\lambda_{2i}PC2_{i}PC2_{i}^{T}}{\lambda_{1i}+\lambda_{2i}} 
% \end{align}

@@

We use multidimensional scaling (MDS) to visualize relationships between windows
as summarized by the dissimilarity matrix $D$,
that produces a set of $m$ coordinates for each window that
give the arrangement in $m$-dimensional space that best recapitulates the original distance matrix.
For results here, we use $m=2$ to produce one- or two-dimensional visualizations of relationships between windows' population structure.

We then locate variation in population structure along the genome
by choosing collections of windows that are nearby in MDS cooridinates,
and map their positions along the genome.
Their common population structure is identified by extracting the corresponding genomic regions
and performing PCA on all, aggregated regions.

@@

\begin{figure}
    \begin{center}
       \includegraphics[width=0.9\textwidth]{Fig1_allchr_Together_MDS_plot_compact_with_ChrX_inv}
    \end{center}
    \caption{
        Variation in structure for windows across \textit{Drosophila melanogaster} chromosome arms.
        In all plots, each point represents one window along the genome.
         The first column shows the MDS visualization of relationships between windows, 
         and the second and third columns show the midpoint of each window against the two MDS coordinates; 
         rows correspond to chromosome arms.
         Colors are consistent for plots in each row. 
         Vertical lines show the breakpoints of known polymorphic inversions.   
         Solid black lines are for the inversions we used to obtaining Figure \ref{fig:pca_by_inversion},
         while dotted grey lines are for other known inversions.     
         \label{fig:mds_allchr}
    }
\end{figure}

@@

The variation in population structure turns out to be almost entirely explained by
several large inversions that are polymorphic in these samples, identified in \citet{corbett2012population}. 
% Frequencies of each inversion in the sample are given in Table \ref{tab:inversion_freqs}.
We recolored the PCA plots in Figure \ref{fig:pca_by_pop} by the orientation of the inversion for each sample,
obtaining Figure \ref{fig:pca_by_inversion}.
Taking chromosome arm 2L as an example,
the two regions of similar, extreme population structure
shown in green in the first row of Figure \ref{fig:mds_allchr}
lie directly around the breakpoints of the inversion In(2L)t,
and the PCA plots in the first rows of Figure \ref{fig:pca_by_inversion}
shows that population structure here is mostly determined by inversion orientation.
The regions shown in purple on chromosome 2L lie near the centromere,
and have population structure reflective of two axes of variation,
seen in Figure \ref{fig:pca_by_pop},
which correspond roughly to latitude within Africa and to degree of cosmopolitan admixture respectively
(see \citet{lack2015drosophila} for more about admixture in this sample).
The regions shown in orange on chromosome 2L mostly lie inside the inversion,
and show population structure that is a mixture between the other two,
as expected due to recombination within the (long) inversion \citep{guerrero2011coalescent}.
Similar results are found in other chromosome arms,
except that the situation of Chr3R is a little more complicated,
due to the coexistence of two polymorphic inversions (In(3R)K and In(3R)P) at intermediate frequency.
However, peaks in MDS coordinates along the chromosome show the breakpoints of both,
although we have not highlighted portions of the MDS plot that pick these out.
Two more inversions, In(3L)P and In(3R)Mo, are at very low frequency in these samples, 
and do not significantly influence population structure.

\begin{figure}
    \begin{center}
       \includegraphics{Fig3_pca_plots_color_by_inv_allchr_with_ChrX_In1A}
    \end{center}
    \caption{
         As in Figure 2, except that samples are colored by orientation of the corresponding polymorphic inversions, 
         In(2L)t, In(2R)NS, In(3L)OK, In(3R)K and In(1)A respectively. 
         \citep[Data from][.]{lack2015drosophila}
        \label{fig:pca_by_inversion}
    }
\end{figure}

@@

As for the \textit{Drosophila} data, we ran our method separately on all 22 human autosomes, 
with results shown in Supplementary Figure \ref{fig:mds12_chr1_8_human}, \ref{fig:mds12_chr9_16_human}, \ref{fig:mds12_chr17_22_human}.
On each, variation in population structure was dominated by a small number of windows
having similar population structure to each other that differed dramatically from the rest of the chromosome.
These may be primarily inversions:
for instance, the eleven windows that are outliers in the first MDS coordinate of chromosome 8 (Figure \ref{fig:mds_human}) 
coincide with the position of a particularly large, polymorphic inversion on 8p23. 
More generally, we see signals coincident with 
three of the six large polymorphic inversions described in \citet{antonacci2009characterization},
shown in Figure \ref{fig:mds_human}.
PCA plots of these windows show a characteristic trimodal shape (Figure \ref{fig:pca_chr8_outliers}),
presumably distinguishing samples having each of the three diploid genotypes for each inversion orientation
(although we do not have data on orientation status),
a signal that has been proposed as a method to identify inversions \citep{ma2012investigation}.
However, distinguishing this hypothesis from others,
such as regions of low recombination rate,
would require additional data.

We also ran the method on all 22 autosomes together, 
and found that, remarkably, 
the inversion on chromosome 8 is still the most striking outlying signal (Figure \ref{fig:mds1_along_allchr_human}). 
Further investigation with a denser set of SNPs,
allowing a finer genomic resolution,
may yield other patterns.

\begin{figure}
    \begin{center}
       \includegraphics[width=0.9\textwidth]{Fig4_POPRES_Together_MDS_plot_chr8_15_17}
    \end{center}
    \caption{
         Variation in structure between windows on human chromosomes 8, 15, and 17. 
         Each point in each plot represents a window.
         The first column shows the MDS visualization of relationships between windows;
         the second and third columns show the two MDS coordinates of each window against its position (midpoint) along the chromosome. 
         Rows, from top to bottom show chromosomes 8, 15, and 17. 
         The vertical red lines show the breakpoints of known inversions from \citep{antonacci2009characterization}.
        \label{fig:mds_human}
    }
\end{figure}

@@

Unlike the other two species,
the method applied separately on all eight chromosomes of \textit{Medicago truncatula} 
showed similar patterns of gradual change in population structure across each chromosome,
with no indications of chromosome-specific patterns.
This consistency suggests that the factor driving the population structure for each chromosome is the same,
as might be caused by varying strengths of linked selection.
To verify that variation in population structure is shared across chromosomes,
we applied the method to all chromosomes together.
Results for chromosome 3 are shown in Figures \ref{fig:mds12_medicago} and \ref{fig:pca_medicago},
and other chromosomes are similar:
the two types of population structure with higher gene density (colored green and purple in Figure \ref{fig:mds12_medicago}
are nearly identical across all chromosomes,
but population structure associated with low gene density (colored orange)
shows more chromsome-specific patterns (see Supplemental Figure \ref{fig:pca_drosophila_allchr}).

Across chromosomes, the high values of MDS coordinate 1 coincide with the position of the heterochromatic regions surrounding the centromere,
which often have lower gene density and may therefore be less subject to linked selection \citep{kulikova2001integration,paape2013selection}. 
To verify that this is a possible explanation,
we computed gene density for each window using gene models in Mt4.0 JBrowse \citep{tang2014improved}. 
we downloaded the gene model track from Mt4.0 JBrowse at jcvi.org,
extracted the start and end positions of each predicted mRNA transcript,
counted how many of these fell in each window, 
and use the mean number of the starts and ends in a window to represent the gene count for that window.
The endpoints of the windows were defined to be halfway between the last SNP of each window and the first SNP in the next window.
These values are shown juxtaposed with
the first MDS coordinate of each window is shown in Figure \ref{fig:mds_medicago},
and are significantly correlated, as shown in Figure \ref{fig:mds_gene_count}.
Other genomic features covary with gene density, however,
so we cannot rule out alternative hypotheses.

>>>>

The general steps in our method could be carried out in many ways.
As depicted in Figure \ref{fig:diagram}, these are
(1) divide the genome into windows,
(2) summarize the patterns of relatedness (population structure) in each window,
(3) measure dissimilarity in population structure between each pair of windows,
(4) visualize the resulting dissimilarity matrix,
and (5) combine similar windows to more accurately estimate local population structure.
Details of how we carried these out are given below.

@@

We used three publically available datasets (summarized in Table \ref{tab:data_stats});
we converted each to a numeric matrix
(with one row per polymorphic variant and one column per sample)
by replacing each genotype with the number of nonreference alleles
(or NA for missing data).
A normalization step (see below) ensures the result does not depend on the choice of reference allele.

\paragraph{\textit{Drosophila melanogaster}:}
We used whole-genome sequencing data 
from the Drosophila Genome Nexus \citep[\url{http://www.johnpool.net/genomes.html},][]{lack2015drosophila},
consisting of the Drosophila Population Genomics Project phases 1--3 \citep{langley2012genomic,pool2012population},
and additional African genomes \citep{lack2015drosophila}.
After removing 20 genomes with more than 8\% missing data,
we were left with 380 samples from 16 countries across Africa and Europe.
Since the \textit{Drosophila} samples are from inbred lines or haploid embryos, 
we treat the samples as haploid when recoding;
regions with residual heterozygosity were marked as missing in the original dataset;
we also removed positions with more than 20\% missing data. 
% (We chose these cutoffs as the tails of the relevant empirical distributions.)
Each chromosome arm we investigated (X, 2L, 2R, 3L, and 3R) has 2--3 million SNPs;
PCA plots for each arm are shown in Figure \ref{fig:pca_drosophila_allchr}.

\paragraph{Human:}
We also used genomic data from the entire POPRES dataset \citep{nelson2008population},
which has array-derived genotype information for 447,267 SNPs across the 22 autosomes
of 3,965 samples in total: 346 African-Americas, 73 Asians, 3,187 Europeans and 359 Indian Asians.
Since these data derive from genotyping arrays, the SNP density is much lower than the other datasets,
which are each derived from whole genome sequencing.
We excluded the sex chromosomes and the mitochondria.
PCA plots for each chromosome, separately, are shown in Figure \ref{fig:pca_human_allchr}.
% We use the allele that has highest frequency in the samples as the reference allele for each position.

@@

\paragraph{\textit{Medicago truncatula}:}
Finally, we used whole-genome sequencing data from the \textit{Medicago truncatula} Hapmap Project \citep{tang2014improved},
which has 263 samples from 24 countries,
primarily distributed around the Mediterranean basin.
Each of the 8 chromosomes has 3--5 million SNPs;
PCA plots for these are shown in Figure \ref{fig:pca_medicago_allchr}.
we did not use the mitochondria or chloroplasts.

\begin{table}[ht]
\centering
    \begin{tabular}{p{0.8in}rrrr}
  \hline
    species 
    & \parbox[t]{.8in}{\# SNPs per \\ window} 
    & \parbox[t]{1in}{mean window\\ length (bp)}
    & \parbox[t]{1.2in}{mean \# windows \\ per chromosome} 
    & \parbox[t]{1.4in}{mean \% variance ex-\\plained by top 2 PCs} \\ 
  \hline
  \textit{Drosophila melanogaster} & 1,000 & 9,019 & 2,674 & 0.53 \\ 
  Human & 100 & 636,494 & 203 & 0.55 \\ 
  \textit{Medicago truncatula} & 10,000 & 102,580 & 467 & 0.50 \\ 
   \hline
\end{tabular}
\caption{
    Descriptive statistics for each dataset used.
    % with the chosen window sizes.
    % See text for how window sizes were chosen;
    Columns 2--4 give statistics describing the window sizes used for most analyses for each dataset
    (windows had a fixed number of SNPs).
    The final column gives the percent variance explained by the top two principal components,
    averaged across independent PCA of each window.
    \label{tab:data_stats}
}
\end{table}

@@

After recoding, we divided the genome into contiguous segments (``windows''; see below),
% recoded matrix into contigous matrices that have the same columns but fewer rows than the original matrix, 
and applied Principal Component Analysis (PCA) as described in \citet{mcvean2009genealogical}
separately to the submatrices that corresponded to each window.
% Specifically, we did PCA as follows:
Precisely,
denote by $Z$ the $L\times N$ recoded genotype matrix ($L$ is the number of SNPs and $N$ is the sample size), 
and by $\overline{Z_{s}}$ the mean of non-missing entries for allele $s$, 
so that $\overline{Z_{s}}=\frac{1}{n_s}\sum_j Z_{sj}$, 
where the sum is over the $n_s$ nonmissing genotypes.
We first compute the mean-centered matrix $X$, as $X_{si}=Z_{si}-\overline{Z_{s}}$,
and preserving missingness.
(This mean-centering makes the result not depend on the choice of reference allele,
exactly if there is no missing data, and approximately otherwise.)
Next, we find the covariance matrix of $X$, denoted $C$,
as $C_{ij} = \frac{1}{m_{ij}-1} \sum_s X_{si} X_{sj} - \frac{1}{m_{ij}(m_{ij}-1)} (\sum_s X_{si})(\sum_s X_{sj})$,
where all sums are over the $m_{ij}$ sites where both sample $i$ and sample $j$ have nonmissing genotypes.
% We compute the covariance matrix using the R function cov() , with use=``pairwise",
% which computes the covariance between each pair of individuals using all complete pairs of SNPs on those individuals.
The principal components are the eigenvectors of $C$, 
normalized to have Euclidean length equal to one,
and ordered by magnitude of the eigenvalues.

@@

Since eigenvectors are still only defined up to sign,
when comparing between windows we choose signs to best match each other:
after choosing $\pcone_1$, for instance, 
if $u$ is the first eigenvector obtained from the covariance matrix
for window $j$,
then we next choose $\pcone_j = \pm u$,
where the sign is chosen according to which of 
$\| \pcone_{1} - u \|$ or
$\| \pcone_{1} + u \|$ 
is smaller.

@@

The choice of window length entails a balance between signal and noise.
In very short windows, genealogies of the samples will only be represented by a few trees,
so variation between windows represents demographic noise rather than meaningful variation in population structure.
Longer windows generally have more distinct trees (and SNPs), 
allowing for less noisy estimation of local population structure.
However, to better resolve meaningful signal, i.e., differences in population structure along the genome,
we would like reasonably short windows.
% Window length choice therefore entails a signal versus noise tradeoff
% in the estimates of population structure.
% If we use the first principal component as a measure of population structure, 
% then to choose the best window length,
% we need to find a balance between the standard error of the first principal components for each window 
% and the standard deviation between windows.

Since we summarize population structure using relative positions in the principal component maps,
we quantify ``noise'' as the standard error of a sample's position on PC1 in a particular window,
averaged across windows and samples,
and ``signal'' as the standard deviation of the sample's position on PC1 over all windows,
averaged over samples.
(Recall that the signs for PCs are chosen to match each other.)
Then, the mean variance across windows is
\begin{align*}
    \sigma_\text{signal}^2
    = 
    \frac{1}{N} \sum_{j=1}^{N}
        \frac{1}{L}\sum_{i=1}^{L}\left ( \pcone_{ij} -\overline\pcone_{j} \right )^{2} ,
\end{align*}
where $\pcone_{ij}$ is the position of the $i^\text{th}$ individual on $\pcone$ in window $j$,
and $\overline\pcone_j = (1/N) \sum_{j=1}^N \pcone_{ij}$.
We estimate the standard error for each $\pcone_{ij}$ using the block jackknife \citep{efron1982jackknife,busing1999deletem}:
we divide the $j^\text{th}$ window into 10 equal-sized pieces,
and let $\pcone_{ij,k}$ denote the first principal component of this region found after removing the $k^\text{th}$ piece;
then the estimate of the squared standard error is
$\sigma^2_{ij} = \frac{9}{10} \sum_{k=1}^{10} ( \pcone_{ij,k} - \frac{1}{10} \sum_{\ell=1}^{10} \pcone_{ij,\ell} )^2$.
Averaging over samples and windows,
\begin{align*}
    \sigma^2_\text{noise}
    &=
    \frac{1}{N} \sum_{j=1}^{N} \frac{1}{L}\sum_{i=1}^{L} \sigma^2_{ij} .
\end{align*}

For the main analysis, we defined windows to each consist of the same number of neighboring SNPs,
and calculated $\sigma^2_\text{signal}$ and $\sigma^2_\text{noise}$
for a range of window sizes (i.e., numbers of SNPs).
For our main results we
chose the smallest window for which $\sigma^2_\text{signal}$ was consistently larger than $\sigma^2_\text{noise}$ (but checked other sizes);
the values for various window sizes across \textit{Drosophila} chromosomes are shown in Table \ref{tab:window_sizes}.
% and the choices for all taxa are in Table \ref{tab:data_stats}.
% In practice, even substantially different window sizes result in the same genome-wide patterns.
In the cases we examined, we found nearly identical results after varying window size,
and choosing windows to be of the same physical length (in bp) rather than in numbers of SNPs.

@@

We think of local population structure as being summarized by \emph{relative} position of the samples
in the space defined by the top principal components.
Since rotations of this space imply identical population structure,
we do not compare population structure of different genomic windows using the PCs directly, 
but instead compare the low-dimensional approximations of the local covariance matrices
obtained using the top $k$ PCs.
(For results shown here, we use $k=2$;
results using larger numbers of PCs were nearly identical.)
We do this, rather than using the entire covariance matrix, 
both for computational efficiency 
and because the top PCs summarize important population structure, 
so using only these should reduce the effect of noise.
To remove the effect of variation in mutation rate between windows,
we also rescale each covariance matrix so that the underlying data matrix has trace norm equal to one.

To do this, define the $N \times k$ matrix $V(i)$ so that $V(i)_{\cdot \ell}$, 
the $\ell^\text{th}$ column of $V(i)$,
is equal to the $\ell^\text{th}$ princpal component of the $i^\text{th}$ window,
multiplied by $\sqrt{ \lambda_{\ell i} / \sum_{m=1}^k \lambda_{m i} }$,
where $\lambda_{\ell i}$ is the $\ell^\text{th}$ eigenvalue of the genetic covariance matrix.
% \begin{equation}
%     \begin{aligned}
%         V_{i1}&=\sqrt{\frac{\lambda _{1i}}{\lambda _{1i}+\lambda _{2i}}}PC1_{i} ,
%         \qquad
%         V_{i2}&=\sqrt{\frac{\lambda _{2i}}{\lambda _{1i}+\lambda _{2i}}}PC2_{i} 
%     \end{aligned}
% \end{equation}
Then, the rescaled, rank $k$ approximate covariance matrix for the $i^\text{th}$ window is
\begin{align} \label{eqn:est_cov}
    M(i) &= \sum_{\ell=1}^k V(i)_{\cdot \ell} V(i)_{\cdot \ell}^T .
\end{align}
% \begin{align}
%     M_{i} &= \frac{\lambda_{1i}PC1_{i}PC1_{i}^{T}+\lambda_{2i}PC2_{i}PC2_{i}^{T}}{\lambda_{1i}+\lambda_{2i}} 
% \end{align}

@@

We use multidimensional scaling (MDS) to visualize relationships between windows
as summarized by the dissimilarity matrix $D$.
MDS produces a set of $m$ coordinates for each window that
give the arrangement in $m$-dimensional space that best recapitulates the original distance matrix.
For results here, we use $m=2$ to produce one- or two-dimensional visualizations of relationships between windows' population structure.

We then locate variation in population structure along the genome
by choosing collections of windows that are nearby in MDS coordinates,
and map their positions along the genome.
Their common population structure is visualized by extracting the corresponding genomic regions
and performing PCA on all, aggregated, regions.

@@

\begin{figure}
    \begin{center}
       \includegraphics[width=0.9\textwidth]{Fig1_allchr_Together_MDS_plot_compact_with_ChrX_inv}
    \end{center}
    \caption{
        Variation in structure for windows across \textit{Drosophila melanogaster} chromosome arms.
        In all plots, each point represents one window along the genome.
         The first column shows the MDS visualization of relationships between windows, 
         and the second and third columns show the midpoint of each window against the two MDS coordinates; 
         rows correspond to chromosome arms.
         Colors are consistent for plots in each row. 
         Vertical lines show the breakpoints of known polymorphic inversions.   
         Solid black lines are for the inversions we used in Figure \ref{fig:pca_by_inversion},
         while dotted grey lines are for other known inversions.     
         \label{fig:mds_allchr}
    }
\end{figure}

@@

The variation in population structure turns out to be almost entirely explained by
several large inversions that are polymorphic in these samples, 
disussed in \citet{corbett2012population} and \citet{langley2012genomic}.
% Frequencies of each inversion in the sample are given in Table \ref{tab:inversion_freqs}.
To depict this, Figure \ref{fig:pca_by_inversion} shows
the PCA plots in Figure \ref{fig:pca_by_pop} recolored by the orientation of the inversion for each sample.
Taking chromosome arm 2L as an example,
the two regions of similar, extreme population structure
shown in green in the first row of Figure \ref{fig:mds_allchr}
lie directly around the breakpoints of the inversion In(2L)t,
and the PCA plots in the first rows of Figure \ref{fig:pca_by_inversion}
shows that population structure here is mostly determined by inversion orientation.
The regions shown in purple on chromosome 2L lie near the centromere,
and have population structure reflective of two axes of variation,
seen in Figure \ref{fig:pca_by_pop},
which correspond roughly to latitude within Africa and to degree of cosmopolitan admixture respectively
(see \citet{lack2015drosophila} for more about admixture in this sample).
The regions shown in orange on chromosome 2L mostly lie inside the inversion,
and show population structure that is a mixture between the other two,
as expected due to recombination within the (long) inversion \citep{guerrero2011coalescent}.
Similar results are found in other chromosome arms,
albeit complicated by the coexistence of more than one polymorphic inversion;
however, each breakpoint visibly affects patterns in the MDS coordinates.
% except that the situation of Chr3R is a little more complicated,
% due to the coexistence of two polymorphic inversions (In(3R)K and In(3R)P) at intermediate frequency.
% However, peaks in MDS coordinates along the chromosome show the breakpoints of both,
% although we have not highlighted portions of the MDS plot that pick these out.
% The inversions In(3L)P and In(3R)Mo, are at low frequency in these samples, 
% and visibly affect the MDS plots although less than the more common inversion on each arm.
% The very young inversion In(X)be shows little signal 
% compared to the older inversion In(X)A \citep{corbett2012population}.

\begin{figure}
    \begin{center}
       \includegraphics{Fig3_pca_plots_color_by_inv_allchr_with_ChrX_In1A}
    \end{center}
    \caption{
         As in Figure 2, except that samples are colored by orientation of the corresponding polymorphic inversions, 
         In(2L)t, In(2R)NS, In(3L)OK, In(3R)K and In(1)A respectively. 
         \citep[data from][]{lack2015drosophila}
        \label{fig:pca_by_inversion}
    }
\end{figure}

@@

As for the \textit{Drosophila} data, we ran our method separately on all 22 human autosomes.
On each, variation in population structure was dominated by a small number of windows
having similar population structure to each other that differed dramatically from the rest of the chromosome.
These may be primarily inversions: outlying windows coincide with
three of the six large polymorphic inversions described in \citet{antonacci2009characterization},
notably a particularly large, polymorphic inversion on 8p23 (Figure \ref{fig:mds_human}). 
% for instance, the eleven windows that are outliers in the first MDS coordinate of chromosome 8 (Figure \ref{fig:mds_human}) 
Similar plots for all chromosomes are shown
in Supplementary Figures \ref{fig:mds12_chr1_8_human}, \ref{fig:mds12_chr9_16_human}, \ref{fig:mds12_chr17_22_human}.
PCA plots of many outlying windows show a characteristic trimodal shape 
(shown for chromosome 8 in Figure \ref{fig:pca_chr8_outliers}),
presumably distinguishing samples having each of the three diploid genotypes for each inversion orientation
(although we do not have data on orientation status).
This trimodal shape has been proposed as a method to identify inversions \citep{ma2012investigation},
but distinguishing this hypothesis from others,
such as regions of low recombination rate,
would require additional data.

We also ran the method grouping all 22 autosomes together, 
and found that, remarkably, 
the inversion on chromosome 8 is still the most striking outlying signal (Figure \ref{fig:mds1_along_allchr_human}). 
Further investigation with a denser set of SNPs,
allowing a finer genomic resolution,
may yield other patterns.

\begin{figure}
    \begin{center}
       \includegraphics[width=0.9\textwidth]{Fig4_POPRES_Together_MDS_plot_chr8_15_17}
    \end{center}
    \caption{
         Variation in structure between windows on human chromosomes 8, 15, and 17. 
         Each point in each plot represents a window.
         The first column shows the MDS visualization of relationships between windows;
         the second and third columns show the two MDS coordinates of each window against its position (midpoint) along the chromosome. 
         Rows, from top to bottom show chromosomes 8, 15, and 17. 
         The vertical red lines show the breakpoints of known inversions from \citet{antonacci2009characterization}.
        \label{fig:mds_human}
    }
\end{figure}

@@

Unlike the other two species,
the method applied separately on all eight chromosomes of \textit{Medicago truncatula} 
showed similar patterns of gradual change in population structure across each chromosome,
with no indications of chromosome-specific patterns.
This consistency suggests that the factor driving the population structure for each chromosome is the same,
as might be caused by varying strengths of linked selection.
To verify that variation in population structure is shared across chromosomes,
we applied the method to all chromosomes together.
Results for chromosome 3 are shown in Figures \ref{fig:mds12_medicago} and \ref{fig:pca_medicago},
and other chromosomes are similar:
% the two types of population structure with higher gene density 
% (colored green and purple in Figure \ref{fig:mds12_medicago})
% are nearly identical across all chromosomes,
% but population structure associated with low gene density (colored orange)
% shows more chromsome-specific patterns (see Supplemental Figure \ref{fig:pca_drosophila_allchr}).
across chromosomes, the high values of the first MDS coordinate coincide with the position of the heterochromatic regions surrounding the centromere,
which often have lower gene density and may therefore be less subject to linked selection \citep{kulikova2001integration,paape2013selection}. 
To verify that this is a possible explanation,
we computed gene densities in each window using gene models in Mt4.0 from \url{jcvi.org} \citep{tang2014improved},
which are shown juxtaposed with
the first MDS coordinate of each window is shown in Figure \ref{fig:mds_medicago},
and are significantly correlated, as shown in Figure \ref{fig:mds_gene_count}.
(Values shown are the number of start and end positions of each predicted mRNA transcript,
divided by two, assigned to the nearest window.)
% that lie closer to SNPs  between the last SNP of each window and the first SNP in the next window.
Other genomic features covary with gene density, however,
so we cannot rule out alternative hypotheses.
